---
title: "Reproducible Research Week 2"
author: "Meri"
date: "10/07/2020"
output: 
  html_document: 
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev ='png')
library(dplyr)
library(ggplot2)
library(lattice) 
```

##Loading and preparing the data

```{r prepping data}
unzip("Activity Monitoring Data.zip")
mydata <- read.csv("activity.csv", header=TRUE)
summary(mydata)
```

##Task 1: What is the mean total number of steps taken per day?

Firstly, caclulate the total number of steps taken per day
```{r total steps}
StepsEachDay <- aggregate(mydata$steps, list(mydata$date), FUN=sum)
colnames(StepsEachDay) <- c("Date", "Steps")
```

Following this, create a histogram of the total number of steps per day
```{r histogram}
g <- ggplot(StepsEachDay, aes(Steps))
g+geom_histogram(fill="light blue", col="dark blue", binwidth = 2500)+ggtitle("Total Number of Steps Per Day")+xlab("Steps")+ylab("Frequency")
```

Lastly, calculate the mean and median values of total number of steps each day
```{r mean}
mean(StepsEachDay$Steps, na.rm=TRUE)
```

```{r median}
median(StepsEachDay$Steps, na.rm=TRUE)
```

##Task 2: What is the average daily activity pattern?

Create a time series plot of 5-minute interval (x axis) and average number of steps taken each day, across all days (y axis)
```{r time series}
StepsPerInterval <- aggregate(data=mydata, steps~interval, FUN=mean)
colnames(StepsPerInterval) <- c("Interval", "Steps")
with(StepsPerInterval, plot(Steps, type="l", main="Average Number of Steps per 5-minute Interval", xlab = "Time", ylab="Steps", col="red"))
```

Following this, calculate the time interval which contains the maximum number of steps
```{r max steps}
StepsPerInterval %>% select(Interval, Steps) %>% filter(Steps==max(Steps))
```

##Task 3: Imputing missing values

Calculate how many missing values are there in the entire dataset
```{r total NAs}
sum(is.na(mydata))
```

Then, devise a strategy to fill in the missing values in the dataset. In this case, I will be filling the NA values with mean number of steps for that day
```{r NA replace}
mydata$newsteps <- ifelse(is.na(mydata$steps), mean(mydata$steps, na.rm=TRUE), mydata$steps)
```

Create a new dataset which is equal to the original dataset, but with the missing values filled in
```{r new dataset}
new_mydata <- data.frame(steps=mydata$newsteps, date=mydata$date, interval=mydata$interval)
head(new_mydata, n=6)
```

Make a histogram of the total number of steps per day and calculate and report the mean and median number of steps per day
```{r histogram 2}
NewStepsEachDay <- aggregate(new_mydata$steps, list(new_mydata$date), FUN=sum)
colnames(NewStepsEachDay) <- c("Date", "Steps")

g <- ggplot(NewStepsEachDay, aes(Steps))
g+geom_histogram(fill="light blue", col="dark blue", binwidth = 2500)+ggtitle("Total Number of Steps Per Day")+xlab("Steps")+ylab("Frequency")
```

Lastly, I calculated the mean and median values of total number of steps each day, for this new dataset
```{r new mean}
mean(NewStepsEachDay$Steps)
```

```{r new median}
median(NewStepsEachDay$Steps)
```
Replacing the missing values with the mean number of steps for that day has little change on the mean and median values calculated. However, the frequency of steps has increased in the new dataset, with replaced missing values. It has gone from a frequency of 17 to 25. 

##Task 4: Are there any differences in activity patterns between weekdays and weekends?

Firstly, create a new factor variable with two levels indicating 'weekday' and 'weekend'
```{r new variables}
new_mydata$date <- as.Date(strptime(new_mydata$date, format="%Y-%m-%d"))
new_mydata$day <- weekdays(new_mydata$date)

for (i in 1:nrow(new_mydata)) {
  if (new_mydata[i,]$day %in% c("Saturday","Sunday")) {
    new_mydata[i,]$day<-"weekend"
  }
  else{
    new_mydata[i,]$day<-"weekday"
  }}
head(new_mydata, n=6)
```

Then, form a panel plot of 5-minute time interval (x axis) and average number of steps taken, averaged across all weekdays and weekends (y axis)
```{r panel plot}
NewStepsPerInterval <- aggregate(new_mydata$steps ~ new_mydata$interval + new_mydata$day, FUN=mean)
colnames(NewStepsPerInterval) <- c("Interval", "Day", "Steps")

xyplot(Steps ~ Interval | Day, NewStepsPerInterval, type="l", layout=c(1,2), xlab="Interval", ylab="Steps")
```